EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces

### 2018-11-07 BrainScience/response.letter.3.tex

\documentclass[journal,onecolumn,12pt]{IEEEtran}

\usepackage{amsmath,amssymb,bm}
\usepackage{amsthm, amsfonts}	
\usepackage{bm,bbm}
\usepackage[normalem]{ulem}
\usepackage{color}
\usepackage{fancybox}
\usepackage{url,booktabs}
\usepackage[round]{natbib}

\usepackage{xr}

\title{Reply to Reviewer's Comments on\\
''EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces''}
\author{}

\begin{document}

\maketitle
\pagenumbering{roman}
\setcounter{page}{1}

We are grateful to the reviewer for pointing out relevant issues in our manuscript.

In the following, we discuss how we dealt with each raised issue.

\vskip+1ex
\noindent \dotfill

\section*{\fbox{Reviewer \#2 Round 2 - Transcript:}}

The authors have addressed many of my concerns regarding this manuscript. There are however a few points that could benefit from clarification.

1) While the authors have addressed the issue of having just a single null-EEG stream from a single subject, it appears to be the case still that a single P300 ERP template is being used. This would seem to either assuming that the P300 is generally invariant across trials (which is unlikely) or that the result of this assessment is restricted to the case in which the ERP response of interest does not vary between trials. This should at least be discussed as a limitation in the conclusion section of the manuscript.

2) The write-up on the active modality section is currently slightly vague. For instance, the reader should not be required to consult Riccio et al. 2013 to obtain information on what "feedback" constitutes. Also, perhaps I am missing something, but I am slightly confused as to why pseudo-real datasets involving superimposition of a selected P300 waveform template are being constructed here. If this is a replication of Riccio, doesn't the original stream itself contains naturally occurring P300s which the authors could contrast the detection performance of their selected classification methods of choice on?

3) As indicated in my initial summary section, the value of this work to the clinical community remains unclear. The value of BCI in clinical work has a long and well-established history (e.g., Neuper et al. 2003). The reference cited in support of this in the response document (Chavarriaga et al., 2017) focuses on the challenges of long-term independent BCI use by intended end users which the current study does not seem to focus on.

\subsection*{\ovalbox{General Comments}}
The authors have addressed many of my concerns regarding this manuscript. There are however a few points that could benefit from clarification. 
\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We thank the Reviewer for

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex
1) While the authors have addressed the issue of having just a single null-EEG stream from a single subject, it appears to be the case still that a single P300 ERP template is being used. This would seem to either assuming that the P300 is generally invariant across trials (which is unlikely) or that the result of this assessment is restricted to the case in which the ERP response of interest does not vary between trials. This should at least be discussed as a limitation in the conclusion section of the manuscript. 
\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We apologize for this mistake.  We definitely selected an incorrect wording because the message that we aimed to convey was exactly the opposite.  We have modified the abstract to emphasize the message and the idea that we failed to transmit in our original manuscript.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

2) The write-up on the active modality section is currently slightly vague. For instance, the reader should not be required to consult Riccio et al. 2013 to obtain information on what "feedback" constitutes. Also, perhaps I am missing something, but I am slightly confused as to why pseudo-real datasets involving superimposition of a selected P300 waveform template are being constructed here. If this is a replication of Riccio, doesn't the original stream itself contains naturally occurring P300s which the authors could contrast the detection performance of their selected classification methods of choice on?

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
These important references were added to the manuscript.  Thank you for pointing them out.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

3) As indicated in my initial summary section, the value of this work to the clinical community remains unclear. The value of BCI in clinical work has a long and well-established history (e.g., Neuper et al. 2003). The reference cited in support of this in the response document (Chavarriaga et al., 2017) focuses on the challenges of long-term independent BCI use by intended end users which the current study does not seem to focus on.

\vskip+1ex
\noindent \dotfill
\vskip+1ex

\bibliographystyle{mdpi}
\bibliography{brainscience}

\end{document}

### 2018-11-07 BrainScience/response.letter.3.tex

\title{Reply to Reviewer's Comments on\\
''EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces''}
\author{}

\begin{document}

\maketitle
\pagenumbering{roman}
\setcounter{page}{1}

We are grateful to the reviewer for giving us very valuable information about issues in our manuscript.

@@

2) The write-up on the active modality section is currently slightly vague. For instance, the reader should not be required to consult Riccio et al. 2013 to obtain information on what "feedback" constitutes. Also, perhaps I am missing something, but I am slightly confused as to why pseudo-real datasets involving superimposition of a selected P300 waveform template are being constructed here. If this is a replication of Riccio, doesn't the original stream itself contains naturally occurring P300s which the authors could contrast the detection performance of their selected classification methods of choice on?

3) As indicated in my initial summary section, the value of this work to the clinical community remains unclear. The value of BCI in clinical work has a long and well-established history (e.g., Neuper et al. 2003). The reference cited in support of this in the response document (Chavarriaga et al., 2017) focuses on the challenges of long-term independent BCI use by intended end users which the current study does not seem to focus on.

\subsection*{\ovalbox{General Comments}}
The authors have addressed many of my concerns regarding this manuscript. There are however a few points that could benefit from clarification. 
\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We thank again the Reviewer for the detailed information.

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex
1) While the authors have addressed the issue of having just a single null-EEG stream from a single subject, it appears to be the case still that a single P300 ERP template is being used. This would seem to either assuming that the P300 is generally invariant across trials (which is unlikely) or that the result of this assessment is restricted to the case in which the ERP response of interest does not vary between trials. This should at least be discussed as a limitation in the conclusion section of the manuscript. 
\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We have modified the experiments to use all the available templates.  On the passive-modality, we are using 70 averaged templates obtained from Subject Number 8 of the BNCI-Horizon dataset (Riccio et al).  They are randomly selected and they are superimposed on the EEG stream.

For the active-modality, first all 70 templates are extracted from each subject on an offline averaging procedure.  These 70 templates are superimposed to a random non-hit preprocessed segment (see next issue).

We agree with the Reviewer that our initial approach was naively optimistic in terms of adding just one template to the whole set was naively optimistic.

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

2) The write-up on the active modality section is currently slightly vague. For instance, the reader should not be required to consult Riccio et al. 2013 to obtain information on what "feedback" constitutes. Also, perhaps I am missing something, but I am slightly confused as to why pseudo-real datasets involving superimposition of a selected P300 waveform template are being constructed here. If this is a replication of Riccio, doesn't the original stream itself contains naturally occurring P300s which the authors could contrast the detection performance of their selected classification methods of choice on?

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We have modified the Section 2.6 to clarify in detail the procedure.  We are using a pseudo-real dataset to have certainty of the existence of the P300 response, which is generating by a subject and we are not sure if the response is there or not.

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

3) As indicated in my initial summary section, the value of this work to the clinical community remains unclear. The value of BCI in clinical work has a long and well-established history (e.g., Neuper et al. 2003). The reference cited in support of this in the response document (Chavarriaga et al., 2017) focuses on the challenges of long-term independent BCI use by intended end users which the current study does not seem to focus on.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you very much for this comment.  It pointed out something that we stated incorrectly and which required more thoughtful write-up. We refer to the approach suggested by Chavarriaga et al 2017 is the UCD, User-Centered Design, where they state that "close interaction between family caregivers and medical staff tremendously facilitates the porocess of transferring an AT to the end-user's home as not only can the needs and requirements of the end user can be taken into account but also the device can be adjusted to the needs and requiremens twho support the end user.  A multidisciplinary approach would allow for improvement of end-user training, for example, by taking into account expert opinion on how to create learning paradigms that promote success but also facilitate BCI use for the end-user." (page 4).

The message that we aimed to transmit, and that we hope this new version of the manuscript now incorporates, is that a waveform-based BCI system, which has an inherent intelligible property (REF), will foster a collaboration in a multidisciplinary environment and will ease the acceptance, and the usability and usefulness of a BCI device.   The reason being, for caregivers and medical staff, particularly those with the traditional knowledge of the clinical EEG which is based on waveforms, they will feel more natural the understanding of how the system is detecting signals.

We have modified the 2.6 section and the Conclusions.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

### 2018-11-07 BrainScience/brainscience.tex
Adding modifications per Reviewers comments

@@

\noindent with $k$ representing the index of the selected atom $g_{k}$ and $\left\lVert \cdot \right\rVert$ its euclidean norm.  Finally the contribution of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}

@@

This sequence is of order $m$, which is the number of sample points used to derive the ordinal element called $\pi$. There are $m!$ ways in which this sequence can be orderly arranged, according to the position that each sample point holds within the sequence in a strict order relationship.  For example if $m=3$, and the first sample point is the bigger, the second is the smaller and the third one is in the middle, the ordinal element $\pi$ corresponds to $(3,1,2)$. Thus, along the signal window there can be at most $k$ different ordinal (and overlapping) elements $\pi_{s}$

@@

@@

The experiments are as follows:

@@

BCI researcers seem to ignore the fact that these systems consist of a closed loop between the human and the mahine.  the human is typically considered as a static compliant entity that always performs what the experimenters asks for in a precise and consistent way.

@@

\begin{figure}[htb]
\centering
\subfigure[EEG trace of the original signal.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
\subfigure[The same eight-channel signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
\subfigure[EEG sample of Cz and L channel of the original EEG trace.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
\subfigure[The same segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
\caption{Eight-channel EEG signal for Subject Number 1 of the psuedo-real dataset without and with the superimposed ERP Template. The channel L, the mark which identifies where to superimpose the P300 ERP, is shown as well as the channel S which identifies the stimulus that was presented. On (c) and (d) the small variation that was introduced by the superimposition of the ERP can be seen enclosed by the vertical bars, where the slope of the bump on subfigure (d) is slightly bigger}
\label{fig:gains}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=1.0\linewidth]{images/GainCheck.eps}
\caption{Point-to-point averaged signals.  These are extracted from the first letter identification trial of the Subject 1 of the psuedo-real dataset.  The ERP is superimposed on classes 3 and 9.  Class 3 is obtained while averaging the segments where the row of the speller matrix is intensified whereas class 9 is calculated from the intensification of the corresponding column.}
\label{fig:gaincheck}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{images/dictionary.eps}
\caption{Coherently averaged signals containing the superimposed ERP.  Each one is extracted from the EEG signal of the Subject 1 of the pseudo-real dataset.  These averaged signals correspond to the 15 first trials (2 signals from each trial, one belonging to the column flashing and the other to row flashing).  These are the templates used to build a dictionary per channel per subject and are used by the classification algorithm described in \ref{section:classification}.}
\label{fig:dictionaryfig}
\end{figure}

@@

Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Kathner2017}.  When the character identification rate reaches this level of performance, the usage of word predicting algorithms allows to implement practical speller applications.  Results for this competition have shown perfect classification with tailored algorithm.  This level is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.  It is important to remark that the algorithms presented here analyze the waveform structure of a single-channel signal~\citep{Rakotomamonjy2008,Gribonval2008}.

@@

>>>>

\subfigcapmargin = 4pt

@@

\noindent with $k$ representing the index of the selected atom $g_{k}$ and ${\left\lVert \cdot \right\rVert}_{2}$ its euclidean norm.  Finally the contribution of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}

@@

This sequence is of order $m$, which is the number of sample points used to derive the ordinal element called $\pi$. There are $m!$ ways in which this sequence can be orderly arranged, according to the position that each sample point holds within the sequence in a decreasing order relationship~\citep{Keller2017}.  For example if $m=3$, and the first sample point is the bigger, the second is the smaller and the third one is in the middle, the ordinal element $\pi$ corresponds to $(1,3,2)$. Thus, along the signal window there can be at most $k$ different ordinal (and overlapping) elements $\pi_{s}$

@@

The problem with any P300 experiment is that users are part of a closed loop with the BCI system and the human is typically considered as a static compliant entity that always performs what the experimenter asks for in a precise and consistent way.  Therefore, P300 experiments are often mined with null-signals.  These are EEG streams which are marked as hit but, because the subject was not particularly focused, or concentrated or so on, the real signal is not generated.   This lack of certainty may be in detriment of analysis of algorithms and can be misleading.   Similar experiments had address this same issue, particularly when benchmarking different algorithms.

@@

\subsubsection{Experiments}
The experiments are as follows:

@@

@@

\begin{figure}[htb]
\centering
\subfigure[EEG trace of the original signal. The horizontal axis represents $10$ seconds of the EEG stream, from the $28$th second up to the $38$th.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
\subfigure[The same $10$-seconds eight-channel signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
\subfigure[EEG sample of Cz and L channel of the original EEG trace. Only $1$ second is shown here, at the $31$th second.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
\subfigure[The same segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
\caption{Eight-channel EEG signal for Subject Number 1 of the pseudo-real dataset without and with the superimposed ERP Template. The channel L, the mark which identifies where to superimpose the P300 ERP, is shown as well as the channel S which identifies the stimulus that was presented. On (c) and (d) the small variation that was introduced by the superimposition of the ERP can be seen enclosed by the vertical bars, where the slope of the bump on subfigure (d) is slightly bigger}
\label{fig:gains}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=1.0\linewidth]{images/GainCheck.eps}
\caption{Point-to-point averaged signals.  These are extracted from the first letter identification trial of the Subject 1 of the pseudo-real dataset.  The ERP is superimposed on classes 3 and 9.  Class 3 is obtained while averaging the segments where the row of the speller matrix is intensified whereas class 9 is calculated from the intensification of the corresponding column.}
\label{fig:gaincheck}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{images/dictionary.eps}
\caption{Coherently averaged signals segments of $1$-second length containing the superimposed ERP.  Vertical axis unit is $\mu V$.  Each one is extracted from the EEG signal of the Subject 1 of the pseudo-real dataset.  These averaged signals correspond to the 15 first trials (2 signals from each trial, one belonging to the column flashing and the other to row flashing).  These are the templates used to build a dictionary per channel per subject and are used by the classification algorithm described in \ref{section:classification}.}
\label{fig:dictionaryfig}
\end{figure}

@@

Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Neuper2003,Kathner2017}.  When the character identification rate reaches this level of performance, the usage of word predicting algorithms allows to implement practical speller applications.  Results for this competition have shown perfect classification with tailored algorithm.  This level is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.  It is important to remark that the algorithms presented here analyze the waveform structure of a single-channel signal~\citep{Rakotomamonjy2008,Gribonval2008}.

@@

waveform-based BCI system, which has an inherent intelligible property (REF), will foster a collaboration in a multidisciplinary environment and will ease the acceptance, and the usability and usefulness of a BCI device.   The reason being, for caregivers and medical staff, particularly those with the traditional knowledge of the clinical EEG which is based on waveforms, they will feel more natural the understanding of how the system is detecting signals.
